---
title: "dbi.table"
output: md_document
---

# `dbi.table`

<!-- badges: start -->
[![R-CMD-check](https://github.com/kjellpk/dbi.table/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/kjellpk/dbi.table/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `dbi.table` package allows you to query database tables and views over a
`DBI` connection using `data.table`’s `[i, j, by]` syntax. The package provides
functions for connecting to a single table, for attaching a database schema to
the search path, and for connecting to an entire database catalog. When using
schemas and catalogs, `dbi.table` uses a table's primary key as the default key
when creating locally instantiated `data.table`s and uses a table's foreign key
as the default for `by` when merging.

## Installation

```{r, eval = FALSE}
# Install dbi.table from CRAN:
install.packages("dbi.table")

# Or the development version from GitHub:
# install.packages("remotes")
remotes::install_github("kjellpk/dbi.table")
```

## Quick Start

First, let's load the package.

```{r library, message = FALSE}
library(dbi.table)
library(data.table) #for as.data.table
```

While the `dbi.table` package will work with `DBIConnection` handles (e.g., as
returned by `DBI::dbConnect`), best practice is instead to use a zero argument
function that returns a `DBIConnection`. This paradigm allows the package to
open, close, and reconnect to the database as necessary.

The function `chinook.duckbd` returns an open `DBIConnection` handle to the
sample *Chinook* database included in the package.

```{r dbi.attach}
dbi.attach(chinook.duckdb)
```

The Chinook database is now attached to the search path in position 2.

```{r search_path}
head(search(), 3)
```

Its tables can be queried using `data.table`'s `[i, j, by]` syntax.

```{r first_query}
Track[MediaTypeId == 1, .("#_of_Tracks" = .N), by = .(Composer)]
```

The `csql` utility displays the `dbi.table`'s SQL query.

```{r csql}
csql(Track[MediaTypeId == 1, .("#_of_Tracks" = .N), by = .(Composer)])
```

Simply `detach` the schema when you are finished and the `DBI` connection will
be closed the next time garbage collection runs.

```{r detach}
detach("duckdb:chinook_duckdb")
```

## Supported Databases

As of version 1.0.4, `dbi.table` supports SQLite, DuckDB, Postgres (RPostgres),
and MariaDB (RMariaDB). Unsupported RDBMSs fallback to DBI - not all features
will be available. Full support can be added by implementing a few S3 methods.
Please feel free to open an issue if you would like to use `dbi.table` with a
particular RDBMS.
