---
title: "dbi.table"
output: md_document
---

# `dbi.table`

Query database objects (e.g., tables and views) accessible over a `DBI`
connection using `data.table`-like syntax. Access individual objects using
`dbi.table` or attach an entire schema using `dbi.attach`.


## Installation

```{r, eval = FALSE}
devtools::install_github("kjellpk/dbi.table")
```

## Usage

First, load the package.

```{r, message = FALSE}
library(dbi.table)
```

Open a `DBI` connection to the Chinook database included in the package.

```{r}
db_path <- file.path(system.file(package = "dbi.table"),
                     "example_files",
                     "chinook_sqlite.sqlite")
chinook <- DBI::dbConnect(RSQLite::SQLite(), db_path)
```

Create a `dbi.table` using the `Album` table in the Chinook database.

```{r}
(Album <- dbi.table(chinook, DBI::Id(table = "Album")))
```

The local object `Album` is a `dbi.table`, a data structure that generates and
executes queries on the remote (SQLite) table `Album`. `dbi.table`s use the same
syntax as `data.tables`. For example, the following command selects the
`AlbumId` and `Title` fields from the remote table `Album` where `AlbumId` is
less than 100 and equal to `ArtistId`.

```{r}
(x <- Album[AlbumId < 100 & AlbumId == ArtistId, .(AlbumId, Title)])
```

By default, the preview is the first 5 rows returned by the dbms (in this case
SQLite). There are only 3 records that satify the query so all results were
returned.

The local object `x` is not the data, rather it is another `dbi.table`
corresponding to the specific query. Use empty square brackets (`[]`) or the
`as.data.table` function to fetch the results set as a `data.table`.

```{r}
x[]
```


```{r, include = FALSE, results = "hide"}
DBI::dbDisconnect(chinook)
```
