---
title: "dbi.table"
output: md_document
---

# `dbi.table`

Query database objects (e.g., tables and views) accessible over a `DBI`
connection using `data.table`'s `[i, j, by]` syntax. Under the hood, a
`dbi.table` is a just an SQL query - its methods are designed to give it
`data.table` look and feel. 


## Installation

```{r, eval = FALSE}
devtools::install_github("kjellpk/dbi.table")
```

## Quick Start

First, load the package.

```{r library, message = FALSE}
library(dbi.table)
```

Next, open a DBI connection (big thank you to the [CTU Prague
Relational Learning Repository](https://relational-data.org/) for providing
this example).

```{r connection}
ctu_conn <- DBI::dbConnect(RMariaDB::MariaDB(),
                           host = "relational.fel.cvut.cz",
                           port = 3306,
                           user = "guest",
                           password = "ctu-relational")
```

Then make a `dbi.table` based on the `Album` table in the `Chinook` schema.

```{r dbi.table}
(Album <- dbi.table(ctu_conn, DBI::Id("Chinook", "Album")))
```

A `dbi.table` looks like a `data.table` (but with the row numbers omitted).
Under the hood, a `dbi.table` is a SQL query - the printed output here is
a 5 row preview. This query can be maniulated using `data.table`'s `[i, j, by]`
syntax.

```{r syntax}
(x <- Album[nchar(Title) > 20 & AlbumId > ArtistId,
            .(Title, Title_Length = paste(nchar(Title), "characters"))])
```

Some data wrangling can be done on-database. The `csql` utility displays the
underlying SQL query (SQL generation uses `dbplyr::translate_sql_`).

```{r csqlx}
csql(x)
```

Finally, close the DBI connection.

```{r detach}
DBI::dbDisconnect(ctu_conn)
```
