---
title: "`dbi.table` Technical Details"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true

vignette: >
  %\VignetteIndexEntry{`dbi.table` Technical Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  prompt = TRUE,
  comment = "  ")
```

```{r setup, include = FALSE, output = "hide"}
library(dbi.table)
dbi.attach(chinook.sqlite)
```

# Introduction

This vignette describes the internal workings of the `dbi.table` package. It is
not intended for end-users of the package. Sadly, I (package author) have very
little time for package development these days so I need some notes to remind
myself how my package works.


# The `dbi.table` Object

A `dbi.table` is a list of *language objects* with the class attribute set to
`dbi.table` (i.e., an S3 object). Each language object is either a `call`, a
`name` (single-variable expressions), or a scalar constant (a length-1 vector
with mode numeric, logical, or character).

```{r element}
c(Genre) #use `c` for the side effect of stripping off the attributes
```

The names of the `dbi.table` object correspond to the column names of the
results set and, since `Genre` is a *pristine* (i.e., corresponds exactly to a
database object) `dbi.table`, each language object is a name.

```{r element.mode}
sapply(Genre, mode)
```

See the `fields` attribute for the intepretation of these names.


## Attributes

In addition to `class` and `names`, a `dbi.table` object uses several additional
attributes to store meta-information.

```{r attributes}
names(attributes(Genre))
```

### The `conn` Attribute

The `conn` attribute can either be a `DBIConnection` or an `environment`. When
the `conn` attribute is an `environment`, `conn$.dbi_connection` will be used as
the `DBIConnection`.


### The `data_source` Attribute

The `data_source` attribute stores the information needed to generate the `FROM`
clause. The `data_source` attribute is a 4-column `data.frame`.

| Column  | Description                                                        |
|---------|--------------------------------------------------------------------|
| clause  | `character` - the name of the clause (`FROM`, `INNER JOIN` , etc.) |
| id      | `DBI::Id` - reference to a database object (table or view)         |
| id_name | `character` - the key used to identify the `Id` in `id`            |
| on      | `call` - the `ON` part of a join clause                            |

```{r data_source}
attr(Genre, "data_source")
```

Join operations add a row to the `data_source`.


### The `fields` Attribute

The `fields` attributes stores information about the database fields (columns)
that are in the `data_source`. The `fields` attribute is a 3-column
`data.frame`.

| Column        | Description                                                  |
|---------------|--------------------------------------------------------------|
| internal_name | `character` - identifier used internally by `dbi.table`      |
| id            | `character` - key to row in `data_source`                    |
| field         | `character` - column name in the connected database          |

```{r fields}
attr(Genre, "fields")
```


### The `distinct` Attribute

A logical value. Default `FALSE`. When `TRUE`, the `DISTINCT` keyword is added
to the select clause of the SQL query.


### The `where` Attribute

A list of calls. Any `name`'s in these calls must be in the `internal_name`
column of the `fields` attribute. These calls are translated to SQL then
`AND`'d together to form the `WHERE` clause of the SQL query.


### The `group_by` Attribute

A list of `name`s and `call`s. Any `name`'s in this list's elements must be in
the `internal_name` column of the `fields` attribute. These calls are translated
to SQL to form the `GROUP BY` clause of the SQL query.


### The `order_by` Attribute

A list of `name`s and `call`s. Any `name`'s in this list's elements must be in
the `internal_name` column of the `fields` attribute. The list elements are
translated to SQL to form the `ORDER BY` clause of the SQL query.


### The `ctes` Attribute

A list of `dbi.tables` (each with the `order_by` and `ctes` attributes reset to
empty `list`s). The `dbi.tables` in `ctes` are rendered as SQL to form the
`WITH` clause of the SQL query.


# Manipulating `dbi.table`s

The `sub_lang` function (not exported by the package) substitutes language
elements into a language element. The idea is to manipulate the columns of a
`dbi.table` while preserving the property that all `name`s in the language
elements are in the `fields` attribute.

The function is defined as follows.

```{r expose_unexported, include = FALSE, output = "hide"}
session <- dbi.table:::session
sub_lang <- dbi.table:::sub_lang
```

```{r sub_lang_args}
args(sub_lang)
```

The arguements are styled after `base::eval` but include an additional frame to
support special symbols and functions (e.g, `.` and `.N`). In most cases,
`envir` is a `dbi.table` and `enclos` is the calling frame.

```{r sub_lang_one}
Genre$GenreId
(j <- sub_lang(list(GenreIdp1 = quote(GenreId + 1)), envir = Genre))
```

A further manipulation can create a more complex `call`.

```{r sub_lang_two}
sub_lang(list(GenreIdp1sq = quote(GenreIdp1^2)), envir = j)
```

By default, `sub_lang` will substitute symbols in `enclos` that evaluate to a
scalar (length one atomic vector) value with mode numeric, character, or
logical.

Note: when `sub_lang` encounters a numeric vector containing only integer
values, the storage mode of the vector is changed to `integer`.


# Connection Management

The first idea was to use the `pool` package for connection management. However,
`pool` uses `DBI::dbIsValid` to test whether a connection is *valid*. It turns
out that `DBI::dbIsValid` returns *valid* as long as the connection has not been
specifically closed with `DBI::dbDisconnect`. If a connection is *borked* for
pretty much any other reason, `pool` will still continue to try to use it.

A secondary concern: in the future the `dbi.table` package will need to use
temporary tables - these will only be available with the connection used to
create them.

When using `dbi.table` in interactive mode or programming mode, the connection
may be specified as a zero-arguement function that returns a connection. In this
case, the function is evlauted to obtain a connection and then the function is
saved as an attribute of the connection. When an error is encountered, 
`dbi.table` attemps the query `SELECT 1`. If that also results in an error, the
function is called to reestablish the connection.

```{r cleanup, include = FALSE, output = "hide"}
# Detach so that the finalizer closes the DBI connection
detach("RSQLite:Chinook_Sqlite")
```

