---
title: "`dbi.table` Syntax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`dbi.table` Syntax}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  prompt = TRUE,
  comment = "  ")
options(width = 104)
```

This vignette describes the `data.table` syntax supported in `dbi.table`.

```{r setup, include = FALSE, output = "hide"}
library(dbi.table)
dbi.attach(chinook.sqlite)
```

```{r objects}
class(DB <- Album)
class(DT <- as.data.table(DB))
names(DT)
```

### Supported syntax

Definition: Supported Syntax is where `data.table` does what you want and
doesn't complain.

##### `i` an (integer) numeric vector 

```{r i_numeric}
i <- c(5, 34, 127)
DT[i]
```

```{r cleanup_i, include = FALSE, output = "hide"}
rm(i)
```

Not supported in `dbi.table` because some (most?) RDMSs do not store records in
any particular (i.e., guaranteed reproducible) order. Also, would require
sending `i` (which could potentially be quite large) as part of the query.


##### `i` a logical `call` 

A logical call is the preferred way to subset the rows of a `dbi.table`.

```{r i_logical_call}
DT[AlbumId > 1 & ArtistId < 5]
```

In this case, `i` becomes the `WHERE` clause in the `dbi.table`'s SQL query.

```{r i_logical_call_sql}
csql(DB[AlbumId > 1 & ArtistId < 5])
```

##### `i` a `call` to `order`

```{r i_call_to_order}
DT[order(ArtistId)]
```


##### `j` a `list` of `language` statements

Here, `language` statement means a `call`, a `name`, or a scalar atomic -
things that one would normally expect to find in a parsed statement.  The
`name`s are intended to be column names in the `dbi.table` but some very
selective partial evalutaion is done for convenience.

```{r j_lang_list}
DT[, list(AlbumId, Short_Title = substring(Title, 1, 20), ArtistId)]
```

`data.tables`'s `.` syntax is also supported; `.(args)` is the same as
`list(args)`.

In this case, `j` becomes the `SELECT` clause in the `dbi.table`'s SQL query.

```{r j_lang_list_sql}
csql(DB[, list(AlbumId, Short_Title = substring(Title, 1, 20), ArtistId)])
```

This is the preferred way to select and modify the columns of a `dbi.table`.


##### `j` a single `name`

```{r j_single_name}
DT[, AlbumId]
```

Since `[.dbi.table` always returns a `dbi.table` (except for the `[]` special
case), this supported `data.table` behavior is not supported in `dbi.table`.


##### `j` a no-`name` `call`

In this case `j` is a `call` where `all.vars(j)` gives `character(0)`.

If `j` evaluates to a character vector whose elements are all found among the
column names of the `data.table`, then this subset of the columns is returned as
a `data.table`.

```{r j_no_name_call_char}
DT[, "AlbumId"]
```


If `j` evaluates to a numeric vector whose elements are all in the set
`{1, ..., n}` where `n` is the number of columns in the `data.table`, then this
subset of the columns is returned as a `data.table`.

```{r j_no_name_call_num}
DT[, 1]
```



##### `j` using `base::subset`'s non standard evaluation of `:`



```{r detach}
detach("RSQLite:chinook_sqlite")
```

